---
title: "Reproducible Research - Peer Assessment 1"
author: "BMCB"
date: "October 19, 2014"
output: html_document
---

This Rmarkdown document represents my submission for the Reproducible Research Coursera class's peer assesment 1.

## Loading and preprocessing the data

The data is loaded with read.csv()

```{r}
df <- read.csv("activity.csv", stringsAsFactors=FALSE)

# convert to daily totals
daily_steps_df <- aggregate(steps ~ date, data=df, sum)

```

## What is mean total number of steps taken per day?

The following R code chuck calculates the mean and median of the total steps per day, and outpust a histogram showing the total steps for days included in the dataset.

```{r}
mean(daily_steps_df$steps)
median(daily_steps_df$steps)
hist(daily_steps_df$steps, breaks=20, main="Histogram of Steps per Day")
```

## What is the average daily activity pattern

The following R code chunk shows the daily pattern of steps across all 5 minute intervals, averaged across all days.

```{r}
# convert to interval averages
interval_steps_df <- aggregate(steps ~ interval, data=df, mean)

# plot pattern
plot(interval_steps_df, type="l", main="Average Steps Taken during 5-minute Interval During the Day")

# Find max minute
interval_steps_df[which.max(interval_steps_df$steps),]
```

## Imputing missing values

This R code chunk outputs the numer of rows where the steps column is filled with NA.  It then fills in those missing values with the overall mean for the 5-minute interval.  *As it turns out, imputing the data in this fashion does not change the overall averages very much.  The median daily value remains the same, and the mean rises by 1.  In the daily average histogram, there are more days around the average peak, which makes the graph look taller in the middle and shorter towards each extreme.*


```{r}
# Output total rows with steps = NA
sum(is.na(df$steps))

imputed_df <- merge(df, interval_steps_df, by="interval")
names(imputed_df) <- c("interval", "steps", "date", "avg_steps")
imputed_df$imputed_steps <- ifelse(is.na(imputed_df$steps), imputed_df$avg_steps, imputed_df$steps)

# Confirm that there is no longer any missing data
sum(is.na(imputed_df$imputed_steps))

# See if new data differs from original daily averages
# convert to daily totals
daily_imputed_steps_df <- aggregate(imputed_steps ~ date, data=imputed_df, sum)

mean(daily_imputed_steps_df$imputed_steps)
median(daily_imputed_steps_df$imputed_steps)
hist(daily_imputed_steps_df$imputed_steps, breaks=20, main="Histogram of Steps per Day (with Imputed Data)")
```

## Are there differences in activity patterns between weekdays and weekends?

The following R code chunk categories the imputed data by the day of the week, and shows that weekends show a different pattern in step activity. 

```{r}

imputed_df$day_of_week <- weekdays(as.Date(imputed_df$date))

# New weekend factor variable
imputed_df$weekend <- as.factor(ifelse(imputed_df$day_of_week %in% c("Saturday", "Sunday"), "weekend", "weekday"))

weekend_df <- aggregate(imputed_steps ~ interval + weekend, data=imputed_df, mean)

# Graph differences
library(ggplot2)
ggplot(weekend_df,aes(interval,imputed_steps)) +geom_line() + facet_grid(.~weekend)
```
